---
title: "Duration of *must* in spoken American English"
author: "Susanne Flach"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 6, fig.height = 4)
```

## Background

This project was an interesting excursion into the analysis of spoken data, that came about almost by accident: **do the two meanings of English *must* differ in their duration and other phonetic properties?** Duration of this kind had not been studied at all, probably due to lack of data, but differences wouldn't be surprising, as research has shown that homophones like *time* and *thyme* differ in duration (usually it's the more frequent and/or more 'basic' meaning that is shorter).[^1]

[^1]: Gahl, Susanne. 2008. Time and thyme are not homophones: The effect of lemma frequency on word durations in spontaneous speech. Language 84(3). 474–496. https://doi.org/10.1353/lan.0.0035. Lohmann, Arne. 2018. Time and thyme are NOT homophones: A closer look at Gahl’s work on the lemma-frequency effect, including a reanalysis. Language 94(2). e180–e190. https://doi.org/10.1353/lan.2018.0032 

For this kind of analysis, we need a fairly large corpus, which comes with the trade-off that the word duration determination is coarse-grained and the audio quality is low (by traditional standards in phonetics). But the NewsScape corpus is a pretty interesting resource for many corpus-linguistics tasks. ([See the RedHenLab for more info.](https://www.redhenlab.org/home))

### The two meanings of *must*

English *must* has two meanings: the deontic meaning refers to what has to be done (*We <u>must</u> be vigilant*) while the epistemic meaning refers to a situation that is probably the case (*He <u>must</u> be home; his car is in the driveway*).[^2]

[^2]: Some authors call 'deontic' the 'root' meaning.

The two meanings occur in very distinct lexico-grammatical environments. For example, if followed by a perfect (*That must <u>have been</u> difficult*), the meaning is nearly always epistemic, whereas if followed by a main verb (*We must <u>clean</u> up this mess*), the meaning is usually deontic. So the linguistic environment is a very reliable cue to the meaning of *must*, and because of the greater likelihood of co-occurrence, will also contribute to acoustic shorting.

### The question(s)

**In this project, I was interested in whether the two meanings also differ in duration, which has theoretical implications for word storage, but potentially also for lexical change in grammar.** In general, durational differences are due to

* Frequency of use in general
* Frequency of co-occurrence with surrounding words (e.g., transitional probabilities)
* Emphasis: putting emphasis on words will likely lead to longer duration
* Speech rate: if the conversation is fast-paced, words will be shorter

The frequency-related properties are related to expectation: words that co-occur often, will be more easily retrieved from memory. Conversely, if a word is more unexpected, more effort will probably go into its articulation. In cases such as *time* or *thyme* and *but* and *butt* there is a noticeable frequency difference by orders of magnitude. No such difference exists between the two meanings of *must*, which are equally frequent. Differences in frequency are due to corpus composition -- deontic *must* is more frequent in formal speech situations, while epistemic *must* is rather characteristic of informal and/or spoken-like discourse. In any case, their frequency differences are usually on the same order of magnitude.

As there was little research to go by in terms of hypotheses, I was
mainly interested in exploratory analysis: do the two meanings differ in
duration? If so, by how much? Are there more general differences between
the two meanings?

## The data

The [NewsScape2016](https://multimodalcorpora.org/web/) corpus houses corpus-linguistically (NLP) processed data from US Network television. The closed-captions were force-aligned with the audio to produce annotations of beginnings of ends of words, from which the words' duration in centiseconds could be determined. 

The data can be extracted via a cql-based query common in corpus linguistics. The data points used in the analysis were semi-manually picked from a list of 20,000 uses of *must*. I then manually checked about 2,000 examples against the original video using [RapidAnnotator](https://www.rapidannotator.org/), in order to remove unusable data (erroneous force-alignment, transcription, music, etc.) and to annotate meaning and emphasis. See [Appendix](#appendix).

### Data preparation

```{r prelims}
# load libraries
library(tidyverse)
library(knitr)
library(extrafont)
library(see)
library(beeswarm)
library(ggstatsplot)
loadfonts(device = "postscript")
theme_set(suzR::theme_suzR())

# set colors
PB = c(rgb(0, 102, 204, maxColorValue=255))
PG = c(rgb(153, 204, 0, maxColorValue=255))
```

### Variables

The following shows 10 random examples of data points to illustrate the data structure; a more complete description of the variables is in the Appendix.

tp.before measures the forward transitional probability: for example, *administration* and *must* co-occur 112 times, and since *must* occurs 181378 times, the probability that *must* follows *administration* is 112/181378 = 0.000617495. Conversely, the probability that *do* follows *must*

```{r data}
# load data
must <- read_csv("https://raw.githubusercontent.com/skeptikantin/must_duration/main/data/must.csv")

## anc: undo the transformations
head(must)

must |> 
  mutate(TP.before = exp(TP.before),
         TP.after = exp(TP.after))

# inspect the data structure
set.seed(12345)
must |> 
  select(L1, Match, R1, Meaning, Emphasis, CS, SR, TP.before, TP.after, T.Freq) |> 
  # random selection of 10 lines:
  sample_n(10) |> 
  kable(digits = 3, caption = "Must data: outcome(s) and predictors")
```

The first columns include the match (*must*) and one word to its left and right. **Meaning** indicates whether it's epistemic or deontic. **Emphasis** is a binary variable that indicates if the *must* use occured in a context of particular stress or emphasis. **CS** is the duration of *must* in 1/100th of a second (**c**enti**s**econds). S

The binary categorical variables **Meaning** (deontic, epistemic) and **Emphasis** (yes, no) were hand-annotated while the datapoints were checked against the actual videos. **Emphasis** expresses whether the use of *must* occurs in an expression that is particularly emphasized (which will naturally effect its duration).

The numerical variables include the (main) outcome, duration, in **CS** (centiseconds, non-transformed)

Manually annotated: **Meaning**: deontic or epistemic **Gender**: male
or female (annotator's judgement) **CS**: duration of *must* in
centiseconds (1/100th of a second) **Duration**: log-transformed CS
**SR**: speech rate (in syllables/seconds), based on 200

## Exploratory data analysis

### Descriptive statistics

Overall, deontic *must* is substantially shorter than epistemic *must*, by raw numbers that is a difference of almost 50ms; epistemic *must* is also a bit less variable. This is a statistically 

```{r class.source = "fold-show"}
must |> 
  group_by(Meaning) |> 
  summarise(mean = mean(CS), sd = sd(CS))
```

```{r}
#theme_set(theme_source_sans())


ggplot(must, aes(Meaning, CS * 1000, fill = Meaning)) +
  geom_violin(color = NA) +
  geom_boxplot(width = .2, fill = "grey90", notch = TRUE) +
  labs(
    title = "Duration by meaning",
    subtitle = "Deontic *must* is substantially shorter than epistemic *must*",
    y = "Duration (ms)", x = NULL
    ) +
  scale_fill_manual(values = c(PB, PG)) +
  theme(
    legend.position = "none",
    )

```

### Multifactorial: correlations

Of course the duration of a word depends on a variety of factors. First, if a word occurs in a context of rapid speed in general, then *must* will also be spoken faster. We measured "speech rate" as number of syllables per second in the context 200 words to the left and right of the occurrence:

```{r}
ggplot(must,
       aes(SR, CS * 1000)) +
  geom_jitter(size = 2, alpha = 0.8, color = PB) +
  geom_smooth(se = FALSE, color = PG) +
  labs(title = "Duration by speech rate and emphasis",
       subtitle = "Faster speech leads to shorter *must* duration",
       x = "Speech rate (syllables/seconds)",
       y = "Duratiaon (ms)") +
  scale_color_manual(values = c(PB, PG))
```

```{r}
ggplot(must, aes(Meaning, CS * 1000, fill = Emphasis)) +
  geom_violinhalf(
    data = must |> filter(Emphasis == "yes"),
    color = NA
    ) +
  geom_violinhalf(
    data = must |> filter(Emphasis == "no"),
    color = NA,
    flip = TRUE) +
  labs(
    title = "Duration by meaning and emphasis",
    subtitle = "Emphasis leads to longer duration",
    y = "Duration (ms)", x = NULL
    ) +
  scale_fill_manual(values = c(PB, PG))
```

```{r}
ggplot(must,
       aes(SR, CS * 1000, group = Emphasis, color = Emphasis)) +
  geom_jitter(size = 2, alpha = 0.8) +
  geom_smooth(se = FALSE) +
  labs(title = "Duration by speech rate and emphasis",
       subtitle = "Faster speech leads to shorter *must* duration",
       x = "Speech rate (syllables/seconds)",
       y = "Duratiaon (ms)") +
  scale_color_manual(values = c(PB, PG))

```


## Appendix {#appendix}

I extracted the data from the NewsScape 2016 corpus, a collection of
video transcripts from US Network television from the year 2016, which
totals about 250 million words of running text. The programs are mostly
from the ABC, CNN, or Fox. Closed-captions, which are obligatory in the
US, are force-aligned with the video, and word duration is automatically
extracted; if you are familiar with NLP or corpus linguistics, then
think about word duration as a token-level annotation that is available
in NewsScape2016, alongside traditional annotation layers like
part-of-speech or lemma. The word duration annotation specifies where in
the audio-video the word begins and where it ends, and we took,
trivially, the difference between the two. ([See the RedHenLab for more
info.](https://www.redhenlab.org/home))

### Cleaning & sampling

I extracted 20,000 uses of *must*, and semi-automatically removed all
instances that are likely "unusable data" (e.g., misaligned text/audio,
missing time stamps, etc.), or for which important metrics would not be
available (more on this later). From the remaining data I randomly
selected 2,000 examples, which I manually checked against the actual
video snippets, and removed dirty data that:

<li>Pauses of more than 500ms before and after *must*</li>

<li>Where no speech rate could be determined, i.e., no 200 usable tokens
left and right</li>

While **Meaning** was rather straight-forward, `Emphasis` was trickier:
emphasis is clearly a non-binary property, especially as a property that
affects word duration. The choice for a

Emphasis was coded by myself and a collaborator, who was unaware of the
goal of the project until after annotation; the data contains his
annotations only.

### Variables

```{r echo = FALSE, results = 'asis'}
library(knitr)
must |> 
  select(-c(CPOS, Left, Match, Right, L1, R1)) |> 
  # random selection of 10 lines:
  sample_n(10) |> 
  kable(caption = "Must data: outcome(s) and predictors")
```

## References
Gahl, Susanne. 2008. *Time* and *thyme* are not homophones: The effect of lemma frequency on word durations in spontaneous speech, *Language* 84: 474-496. [10.1353/lan.0.0035](https://doi.org/10.1353/lan.0.0035)